---
title: "HW2"
output: pdf_document
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(nlme)
```

## Question 1

### Question 1(a)
```{r chunk_data import, warning=FALSE}
load("~/Documents/2023Fall/P8157/P8157/MACS-VL.RData")
data = macsVL
macs = data |> 
  group_by(id)  |> 
  mutate(idd = group_indices())  |> 
  ungroup() 
```


```{r chunk_variable sum}
# number of clusters
length(unique(data$id))
# number of measurements within each cluster
obs = data |> group_by(id) |> summarize(n_obs = n())
summary(obs$n_obs)
# follow-up period
fl = data |> group_by(id) |> mutate(max_mon = max(month)) |> 
  filter(month == max_mon)
summary(fl$max_mon)
# time interval between measurements within each cluster
int = data |> 
  group_by(id) |> 
  mutate(delta_mon = month - lag(month))
summary(int$delta_mon)
# baseline vload
vl = data |> group_by(id) |> summarize(vload = first(vload))
summary(vl$vload)
ggplot(vl, aes(x = vload)) + 
  geom_histogram()
ggplot(vl, aes(x = log(vload))) + 
  geom_histogram()
# cd4+ count
c4 = data |> group_by(id) |> summarize(base_cd4 = first(cd4), last_cd4 = last(cd4)) |> 
  mutate(loss_cd4 = base_cd4 - last_cd4)
summary(c4$loss_cd4)
# spaghetti plot
ggplot(data, aes(x = month, y = cd4, group = id, color = id)) + 
  geom_line() 
```

```{r chunk_2 stage model}
K = 225
# Stage 1
betaMat = data.frame(beta0=rep(NA, K), beta.time=rep(NA, K))
for(k in 1:K) {
  temp.k = macs[macs$idd == k,]
  fit.k = lm(log(cd4) ~ month, data = temp.k)
  betaMat[k, 1:2] = c(fit.k$coef)
}

# Stage 2
data_2 = cbind(vl, betaMat)
model_time = lm(beta.time ~ vload, data = data_2)
summary(model_time)
```
The modeling result indicates that vload is certainly a significant modifier of the rate of decline of CD4+ cell count.

### Question 1(b)
```{r chunk_covariance, warning=FALSE}
data_1 = data |> 
  mutate(halfyr = round(month/6))
fitf = lm(cd4 ~ halfyr, data = data_1) 
resMat = matrix(residuals(fitf), ncol=8, byrow=TRUE) 
# covariance matrix diagonal
sd = round(sqrt(diag(cov(resMat))), 2)
sd = c(266.63, 323.47, 312.31, 299.70, 272.13, 315.27, 286.79, 274.45, 332.57)
sd = c(330.30, 264.27, 272.81, 320.29, 338.98, 288.09, 279.74, 292.83)
# correlation
comat = round(cor(resMat), 2)
# sd and corr matrix:
diag(comat) = sd
comat
```
The month variable was mutated into a half-year variable and then I explored the covariance structure of the data. There isn't evident trend whether the variances change with time, but the correlation does seem to be decaying as a function of time between observations. Thus the auto-regressive correlation structure seems most appropriate here. 



### Question 1(c)
```{r}
data0 = data |> 
  mutate(vload = log(vload))
fit1 = gls(cd4 ~ month*vload, method = "ML", data = data0, corr = corCompSymm(form = ~ 1 | id))
summary(fit1)
fit2 = gls(cd4 ~ month*vload, method = "REML", data = data0, corr = corCompSymm(form = ~ 1 | id))
summary(fit2)
```
Both ML and REML give significant estimations of the effect of both baseline virus load on CD4+ cell count and the influence of baseline virus load on the decline rate of cell count. Generally, keeping baseline virus load fixed, with one unit increase in month, the cell count would decrease by -3.08-0.38log(vload); keeping month fixed, with one unit increase in log(virus load), the cell count would decrease by -35.7-0.38month. The p-value of the interaction term is 0.0286, indicating that under a significance level of 0.05, there is a significant association between baseline viral
load and the rate of decline in CD4+.

### Question 1(d)
```{r}
vl = data0$vload
summary(vl)
min = min(vl)
max = max(vl)
med = median(vl)
mean = mean(vl)
q1 = quantile(vl,0.25)
q3 = quantile(vl,0.75)
breaks = c(min-1, q1, med, q3, max+1)
cats = c("1", "2", "3", "4")

dataj = data0 |> 
  mutate(cats = cut(vload, breaks = breaks, labels = cats, right = FALSE))

fit3 = gls(cd4 ~ month*cats, method = "REML", data = dataj, corr = corCompSymm(form = ~ 1 | id))
summary(fit3)
```
\begin{table}
\caption{Model with ML ($\rho$ = 0.5673)}
\centering
\begin{tabular}{|l|c|c|c|c|}
	\hline
	 & Value & Standard error & t-value & p-value\\
	\hline
    Intercept &1108.10 &91.33&12.1336&0.0000\\
	\hline
    month	&-3.0861	&1.7605	&-1.7530	&0.0798\\
	\hline
    vload	&-35.70	&8.99	&-3.9704 &0.0001\\
	\hline
    month:vload	&-0.38	&0.17	&-2.1902	&0.0286\\
	\hline
\end{tabular}
\end{table}

\begin{table}
\caption{Model with REML ($\rho$ = 0.5693)}
\centering
\begin{tabular}{|l|c|c|c|c|}
	\hline
	 & Value & Standard error & t-value & p-value\\
	\hline
    Intercept &1108.10	&91.56	&12.10	&0.00\\
	\hline
    month	&-3.09	&1.76	&-1.7530	&0.0798\\
	\hline
    vload	&-35.70	&9.016	&-3.9599	&0.0001\\
	\hline
    month:vload	&-0.38	&0.17	&-2.1911	&0.0286\\
	\hline
\end{tabular}
\end{table}

\begin{table}
\caption{Stratified Model with REML ($\rho$ = 0.5731)}
\centering
\begin{tabular}{|l|c|c|c|c|}
	\hline
	 & Value & Standard error & t-value & p-value\\
	\hline
    Intercept &855.50&33.56&25.4896&0.0000\\
	\hline
    month	&-5.50&0.63	&-8.7239	&0.0000\\
	\hline
    cats2	&-103.94&46.89	&-2.2165	&0.0268\\
	\hline
    cats3	&-122.28&47.61	&-2.5683	&0.0103\\
	\hline
    cats4	&-186.24&46.71	&-3.9871	&0.0001\\
	\hline
    month:cats2	&-1.95	&0.89	&-2.1949	&0.0283\\
	\hline
    month:cats3	&-2.57	&0.89	&-2.8987	&0.0038\\
	\hline
    month:cats4	&-1.00	&0.90	&-1.1162	&0.2645\\
	\hline
\end{tabular}
\end{table}